# Basic components of a table and how to incorporate those

```{r setup, include=FALSE, echo=FALSE}
library(gt)
library(dplyr)
```

## Making a **gt** table: the `gt()` function is the entry point

When one provides table data to the `gt()` function, it generates a **gt** table object. This function is the initial step in a typical **gt** workflow. Once you possess the **gt** table object, you have the ability to perform styling transformations before rendering it as a display table of different formats.

## The table header and footer

- `tab_header()` and `opt_align_header()`

It is possible to add a table header to the gt table, which includes a title and even a subtitle. The table header is an optional component of the table that appears above the column labels. You have the flexibility to use Markdown formatting for the header's title and subtitle, allowing for greater customization. Additionally, if you intend to use HTML output for the table, you may utilize HTML in either the title or subtitle.

```{r}
gtcars |>
  dplyr::select(mfr, model, msrp) |>
  dplyr::slice(1:5) |>
  gt() |>
  tab_header(
    title = md("Data listing from **gtcars**"),
    subtitle = md("`gtcars` is an R dataset")
  )
```


The table header that is appended to a **gt** table typically has center alignment for both the title and subtitle elements. However, it is possible to adjust the horizontal alignment of the title and subtitle by utilizing the `align` argument. This function provides a quick and easy means to set the alignment to the left or right. It also serves as a convenient shortcut for `<gt_tbl> |> tab_options(heading.align = <align>)`.

```{r}
exibble |>
  gt(rowname_col = "row", groupname_col = "group") |>
  summary_rows(
    groups = "grp_a",
    columns = c(num, currency),
    fns = c("min", "max")
  ) |>
  grand_summary_rows(
    columns = currency,
    fns = total ~ sum(., na.rm = TRUE)
  ) |>
  tab_source_note(source_note = "This is a source note.") |>
  tab_footnote(
    footnote = "This is a footnote.",
    locations = cells_body(columns = 1, rows = 1)
  ) |>
  tab_header(
    title = "The title of the table",
    subtitle = "The table's subtitle"
  ) |>
  opt_align_table_header(align = "left")
```


- `tab_source_note()`
- `md()` and `html()`


## Grouping together column labels with spanners

- `tab_spanner()`
- `tab_spanner_delim()`


## The stub and row groups

- `tab_row_group()`
- `row_group_order()`
- `tab_stubhead()`
- `tab_stub_indent()`


## Adding footnotes

- `tab_footnote()` and the `cells_*()` location helpers
- `opt_footnote_*()`

