# Extension packages {#sec-extension-packages}

```{r setup, include=FALSE, echo=FALSE}
library(gt)
library(dplyr)
```

The **gt** package has inspired the creation of extension packages that build upon its foundation to solve specialized problems. These packages demonstrate the flexibility of **gt**'s architecture and provide ready-made solutions for common table-making tasks in specific domains. In this chapter, we'll explore **gtsummary**, one of the most impactful extension packages, which provides a streamlined workflow for creating clinical and analytical summary tables.

The **gtsummary** package wraps **gt**'s functionality to provide sensible defaults for its target use cases while still allowing full access to **gt**'s customization capabilities. This means you can use the package to quickly generate professional tables and then further refine them using the **gt** functions you've already learned.

## **gtsummary**

The **gtsummary** package provides an elegant way to create publication-ready summary tables and regression model results. Originally developed for biomedical research, it has become an indispensable tool for anyone who needs to present descriptive statistics or model outputs in a professional format.

### The problem **gtsummary** solves

Creating a proper "Table 1" for a research paper involves many tedious steps. You need to calculate summary statistics for continuous and categorical variables, handle missing data appropriately, compare groups using the right statistical tests, and format everything consistently. Before **gtsummary**, this process typically required hundreds of lines of code and careful attention to formatting details. The **gtsummary** package reduces this to just a few lines while producing tables that meet the exacting standards of medical journals.

The package automatically detects variable types and calculates appropriate descriptive statistics. Continuous variables get medians and interquartile ranges (or means and standard deviations), while categorical variables get counts and percentages. Missing values are tracked and reported. When comparing groups, the package selects appropriate statistical tests based on data characteristics.

### Summarizing data with `tbl_summary()`

The `tbl_summary()` function is the workhorse of **gtsummary**. It takes a data frame and produces a formatted summary table with minimal code. The package includes a trial dataset for demonstrating its capabilities, which contains simulated data from 200 patients receiving chemotherapy treatments.

```{r}
#| label: gtsummary-basic
#| message: false

library(gtsummary)

trial |>
  select(trt, age, grade, response) |>
  tbl_summary()
```

This simple call produces a table with properly formatted statistics, clear labels, and handling of missing values. The `age` variable is summarized with median and interquartile range because **gtsummary** detected it as continuous. The `grade` variable shows counts and percentages because it's categorical. Missing values are reported as `"Unknown"` at the bottom of each variable's section.

The real power of `tbl_summary()` emerges when comparing groups. By specifying a `by` variable, you can split your summary statistics across treatment arms or other groupings. Adding `add_p()` automatically selects and applies appropriate statistical tests for each variable.

```{r}
#| label: gtsummary-comparison

trial |>
  select(trt, age, grade, response) |>
  tbl_summary(
    by = trt,
    missing = "ifany",
    label = list(
      age ~ "Patient Age (years)",
      grade ~ "Tumor Grade",
      response ~ "Tumor Response"
    )
  ) |>
  add_p() |>
  add_overall() |>
  modify_header(label = "**Characteristic**") |>
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Treatment Group**") |>
  bold_labels()
```

This example demonstrates several of **gtsummary**'s customization options. The `label` argument provides custom variable names. The `add_overall()` function adds a column with statistics for all patients combined. The `modify_header()` and `modify_spanning_header()` functions adjust column labels. Finally, `bold_labels()` applies bold formatting to variable names.

### Presenting regression results with `tbl_regression()`

Clinical research frequently involves regression modeling, and **gtsummary** provides `tbl_regression()` to present model results in publication-ready format. The function works with many model types including linear models, logistic regression, Cox proportional hazards models, and mixed effects models.

```{r}
#| label: gtsummary-regression

# Fit a logistic regression model
model <- glm(
  response ~ age + stage + grade,
  data = trial,
  family = binomial
)

# Create a formatted table of results
model |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      age ~ "Patient Age",
      stage ~ "T Stage",
      grade ~ "Tumor Grade"
    )
  ) |>
  bold_labels()
```

The `exponentiate = TRUE` argument transforms coefficients to odds ratios, which is the standard presentation for logistic regression. Reference categories are automatically identified and marked. The package also provides functions like `add_global_p()` to add overall p-values for categorical variables with multiple levels, though these require additional dependencies.

### Combining multiple tables

Research papers often present multiple models side by side or combine different analyses into a single display. The **gtsummary** package provides `tbl_merge()` and `tbl_stack()` for these situations.

```{r}
#| label: gtsummary-merge

# Create two regression tables
model1 <- glm(response ~ age + grade, data = trial, family = binomial)
model2 <- glm(response ~ age + stage, data = trial, family = binomial)

tbl1 <- tbl_regression(model1, exponentiate = TRUE)
tbl2 <- tbl_regression(model2, exponentiate = TRUE)

# Merge them side by side
tbl_merge(
  tbls = list(tbl1, tbl2),
  tab_spanner = c("**Model 1**", "**Model 2**")
)
```

This approach is particularly useful for showing how results change as you add or remove covariates, or for presenting models with different outcomes.

### Converting to **gt** for additional customization

Every **gtsummary** table can be converted to a **gt** object using `as_gt()`, which opens up all of **gt**'s formatting capabilities. This is useful when you need styling options beyond what **gtsummary** provides natively.

```{r}
#| label: gtsummary-to-gt

trial |>
  select(trt, age, marker) |>
  tbl_summary(by = trt) |>
  add_p() |>
  as_gt() |>
  tab_header(
    title = md("**Patient Characteristics by Treatment**"),
    subtitle = "Simulated Clinical Trial Data"
  ) |>
  tab_source_note("Data simulated for demonstration purposes")
```

This workflow demonstrates how **gtsummary** and **gt** complement each other. You get the convenience of **gtsummary**'s automatic calculations and sensible defaults, then add **gt**'s rich formatting options for the final presentation.

As you develop your table-making skills, extension packages like **gtsummary** become valuable tools in your toolkit. They handle common tasks efficiently while remaining flexible enough for customization. In the next chapter, we'll explore how you can create your own **gt** extensions to share solutions with the broader community.

## Summary

This chapter has introduced **gtsummary**, a powerful extension package that builds upon **gt**'s foundation for creating statistical summary tables.

The key capabilities we've explored:

- **gtsummary** transforms statistical analysis into publication-ready tables. `tbl_summary()` creates descriptive statistics tables with automatic variable detection and appropriate statistics. `tbl_regression()` formats model outputs with proper coefficient presentation. `tbl_merge()` and `tbl_stack()` combine multiple tables for comprehensive reporting.
- integration with **gt**: **gtsummary** produces **gt** objects, meaning you can further customize its output using any **gt** function. Apply additional formatting, add footnotes, adjust styling (the full **gt** toolkit remains available).

Extension packages embody a powerful pattern: domain experts identifying common needs and encoding solutions in reusable code. The tables they produce meet professional standards while requiring minimal code from users.

The final chapter shows how you can create your own extensions, building functions and packages that address the specific table-making challenges in your domain.
