# Quick reference {#sec-appendix-reference}

```{r setup, include=FALSE, echo=FALSE}
library(gt)
library(dplyr)
```

This appendix provides a consolidated reference for the **gt** package, including informational functions that display helpful lookup tables, descriptions of all built-in datasets, and a quick guide to the most commonly used helper functions.

## Date and time style references

### `info_date_style()`

Displays all 41 available date formatting styles for use with `fmt_date()` and `fmt_datetime()`.
 
```{r}
#| eval: false
info_date_style()
```

The resulting table shows:

- style number: the numeric identifier for each style
- style name: the string identifier (e.g., `"iso"`, `"month_day_year"`)
- example output: how each style renders a sample date
- flexible: whether the style adapts to locale settings

Use style names (or numbers) in your formatting calls:

```r
fmt_date(columns = date_col, date_style = "month_day_year")
fmt_date(columns = date_col, date_style = 14)  
```

### `info_time_style()`

Displays all available time formatting styles for use with `fmt_time()` and `fmt_datetime()`.

```{r}
#| eval: false
info_time_style()
```

The table shows similar information to date styles: style numbers, names, example output, and locale flexibility.

## Currency reference

### `info_currencies()`

Displays a comprehensive table of all supported currencies, including their codes, symbols, and names.

```{r}
#| eval: false
info_currencies()
```

The table includes:

- currency code: the 3-letter ISO 4217 code (e.g., `"USD"`, `"EUR"`, `"JPY"`)
- currency symbol: the symbol used in formatting (e.g., `$`, `€`, `¥`)
- currency name: the full name of the currency

Use currency codes with `fmt_currency()`:

```r
fmt_currency(columns = price, currency = "EUR")
fmt_currency(columns = price, currency = "JPY")
```

## Locale reference

### `info_locales()`

Displays all supported locales for internationalized formatting. Locales affect number formatting, date formatting, and translated text.

```{r}
#| eval: false
info_locales()
```

The table shows locale codes and their corresponding languages/regions. Use these codes with:

- the `locale` argument in `gt()`
- the `locale` argument in formatting functions
- the `locale` argument in `fmt_spelled_num()`

```r
gt(data, locale = "de")
fmt_number(columns = value, locale = "fr")
fmt_date(columns = date, locale = "ja")
```

## Color palette reference

### `info_paletteer()`

The **paletteer** package provides access to hundreds of color palettes from various R packages. The `info_paletteer()` function displays available palettes that work well with `data_color()`.

```{r}
#| eval: false
info_paletteer()
```

The table groups palettes by their source package and shows color swatches for each palette. Use palette names with `data_color()`:

```r
data_color(
  columns = value,
  palette = "viridis::plasma"
)
```

Palette names follow the format `"package::palette"`.

## Flag reference

### `info_flags()`

Displays all available country flags that can be rendered with `fmt_flag()`, along with their country codes.

```{r}
#| eval: false
info_flags()
```

The table shows:

- country code (2-letter): ISO 3166-1 alpha-2 codes
- country code (3-letter): ISO 3166-1 alpha-3 codes
- country name: full country name
- flag preview: the rendered flag icon

Use these codes with `fmt_flag()`:

```r
fmt_flag(columns = country_code)
```

## Icon reference

### `info_icons()`

Displays all available Font Awesome icons for use with `fmt_icon()`.

```{r}
#| eval: false
info_icons()
```

The table shows icon names and rendered previews. Use icon names in your data or with `fmt_icon()`:

```r
dplyr::tibble(
  category = c("file", "folder", "image"),
  icon_name = c("file-alt", "folder", "image")
) |>
  gt() |>
  fmt_icon(columns = icon_name)
```

## Google Fonts reference

### `info_google_fonts()`

Displays recommended Google Fonts for use with **gt** tables, along with information about how to use them.

```{r}
#| eval: false
info_google_fonts()
```

Use Google Fonts with `opt_table_font()` or `google_font()`:
 
```r
gt(data) |>
  opt_table_font(font = google_font("Roboto"))

gt(data) |>
  tab_style(
    style = cell_text(font = google_font("Fira Code")),
    locations = cells_body(columns = code_col)
  )
```

## Unit conversion reference

### `info_unit_conversions()`

Displays all unit conversions supported by the `unit_conversion()` helper function. This is useful when you need to convert between measurement units.

```{r}
#| eval: false
info_unit_conversions()
```

The table shows:

- unit class: the category of measurement (length, mass, volume, etc.)
- from units: available source units
- to units: available target units
- conversion factors: the multipliers used

Use with `unit_conversion()`:

```r
# Get conversion factor from miles to kilometers
unit_conversion(from = "mi", to = "km")

# Apply in a table
dplyr::tibble(distance_miles = c(10, 25, 100)) |>
  gt() |>
  cols_add(distance_km = distance_miles * unit_conversion("mi", "km")) |>
  fmt_number(columns = everything(), decimals = 1)
```

## Built-in datasets

The **gt** package includes 18 datasets for learning and experimentation. These datasets appear throughout the documentation and examples.

### Small example datasets

These compact datasets are ideal for quick experiments and learning the basics of **gt**.

```{r}
#| echo: false
dplyr::tibble(
  Dataset = c("`exibble`", "`gtcars`", "`pizzaplace`"),
  Description = c("A toy tibble with various data types", "Specifications for 47 luxury/supercars", "A year of pizza sales"),
  Rows = c("8", "47", "49,574"),
  `Use Case` = c("Testing formatters", "Car comparisons", "Sales analysis")
) |>
  gt() |>
  fmt_markdown(columns = Dataset)
```

### Country and population data

These datasets provide demographic information useful for geographic visualizations and population analysis.

```{r}
#| echo: false
dplyr::tibble(
  Dataset = c("`countrypops`", "`towny`", "`peeps`"),
  Description = c("Country populations 1960-2024", "Ontario municipality populations", "Personal information for fictional people"),
  Rows = c("~13,000", "414", "100"),
  `Use Case` = c("Time series, maps", "Canadian demographics", "Contact directories")
) |>
  gt() |>
  fmt_markdown(columns = Dataset)
```

### Scientific and technical data

These datasets contain scientific measurements and constants, suitable for demonstrating technical table formatting.

```{r}
#| echo: false
dplyr::tibble(
  Dataset = c("`sza`", "`constants`", "`reactions`", "`photolysis`", "`nuclides`"),
  Description = c("Solar zenith angles by location/time", "Fundamental physical constants", "Atmospheric reaction rates", "Photolysis rates for organic compounds", "Nuclear data for isotopes"),
  Rows = c("816", "30", "1,344", "1,590", "3,384"),
  `Use Case` = c("Heatmaps", "Scientific tables", "Chemistry data", "Chemistry data", "Nuclear chemistry")
) |>
  gt() |>
  fmt_markdown(columns = Dataset)
```

### Financial and geographic data

These datasets cover financial markets and location-based information for diverse table examples.

```{r}
#| echo: false
dplyr::tibble(
  Dataset = c("`sp500`", "`metro`", "`gibraltar`", "`films`"),
  Description = c("S&P 500 daily data 1950-2015", "Paris Metro stations", "May 2023 Gibraltar weather", "Cannes Film Festival competition"),
  Rows = c("16,607", "308", "744", "1,607"),
  `Use Case` = c("Financial analysis", "Geographic data", "Weather data", "Entertainment data")
) |>
  gt() |>
  fmt_markdown(columns = Dataset)
```

### Clinical trial data

These datasets follow pharmaceutical industry conventions, useful for creating clinical study tables.

```{r}
#| echo: false
dplyr::tibble(
  Dataset = c("`rx_adsl`", "`rx_addv`", "`illness`"),
  Description = c("ADSL-style clinical trial data", "ADDV-style clinical trial data", "Lab tests for illness tracking"),
  Rows = c("182", "910", "39"),
  `Use Case` = c("Clinical tables", "Clinical tables", "Medical data")
) |>
  gt() |>
  fmt_markdown(columns = Dataset)
```

### Using built-in datasets

All datasets are automatically available when you load the **gt** package, requiring no additional data import steps.

```{r}
# View first few rows
head(exibble)

# Use in a table
gtcars |>
  dplyr::select(mfr, model, year, hp, mpg_c) |>
  dplyr::slice(1:5) |>
  gt()
```

## Helper functions quick reference

### Text and content helpers

These functions allow you to include formatted text in table headers, labels, and notes.

```{r}
#| echo: false
dplyr::tibble(
  Function = c("`md()`", "`html()`", "`latex()`"),
  Purpose = c("Interpret text as Markdown", "Interpret text as HTML", "Interpret text as LaTeX"),
  Example = c('`tab_header(title = md("**Bold Title**"))`', '`tab_header(title = html("<em>Styled</em>"))`', '`tab_source_note(latex("\\\\textit{Note}"))`')
) |>
  gt() |>
  fmt_markdown(columns = c(Function, Example))
```

### Unit helpers

These functions specify measurement units for sizing columns and other table elements.

```{r}
#| echo: false
dplyr::tibble(
  Function = c("`px()`", "`pct()`"),
  Purpose = c("Specify pixels", "Specify percentage"),
  Example = c("`cols_width(col ~ px(150))`", "`cols_width(col ~ pct(25))`")
) |>
  gt() |>
  fmt_markdown(columns = c(Function, Example))
```

### Dynamic value helpers

These functions enable data-driven formatting by referencing values from other columns or defining custom formats.

```{r}
#| echo: false
dplyr::tibble(
  Function = c("`from_column()`", "`currency()`"),
  Purpose = c("Reference column values", "Custom currency symbol"),
  Example = c('`fmt_number(decimals = from_column("precision"))`', '`fmt_currency(currency = currency(symbol = "BTC"))`')
) |>
  gt() |>
  fmt_markdown(columns = c(Function, Example))
```

### Color helpers

These functions help manipulate colors for consistent styling across your tables.

```{r}
#| echo: false
dplyr::tibble(
  Function = "`adjust_luminance()`",
  Purpose = "Lighten/darken colors",
  Example = '`adjust_luminance("steelblue", steps = 1)`'
) |>
  gt() |>
  fmt_markdown(columns = c(Function, Example))
```

### Font helpers

These functions provide access to web fonts and system font stacks for typography customization.

```{r}
#| echo: false
dplyr::tibble(
  Function = c("`google_font()`", "`default_fonts()`", "`system_fonts()`"),
  Purpose = c("Specify Google Font", "Get default font stack", "Get system font stack"),
  Example = c('`opt_table_font(font = google_font("Roboto"))`', "`default_fonts()`", '`system_fonts("humanist")`')
) |>
  gt() |>
  fmt_markdown(columns = c(Function, Example))
```

### Location helpers (for `tab_style()`, `tab_footnote()`, etc.)

These functions target specific regions of the table for styling, footnotes, or other modifications.

```{r}
#| echo: false
dplyr::tibble(
  Function = c("`cells_title()`", "`cells_stubhead()`", "`cells_column_labels()`", "`cells_column_spanners()`", "`cells_row_groups()`", "`cells_stub()`", "`cells_body()`", "`cells_summary()`", "`cells_grand_summary()`", "`cells_stub_summary()`", "`cells_stub_grand_summary()`", "`cells_footnotes()`", "`cells_source_notes()`"),
  Target = c("Table title and subtitle", "Stubhead label", "Column labels", "Column spanner labels", "Row group labels", "Stub (row labels)", "Body cells", "Group summary cells", "Grand summary cells", "Stub cells in summaries", "Stub cells in grand summary", "Footnote area", "Source note area")
) |>
  gt() |>
  fmt_markdown(columns = Function)
```

### Style helpers (for `tab_style()`)

These functions define visual properties that can be applied to targeted table locations.

```{r}
#| echo: false
dplyr::tibble(
  Function = c("`cell_text()`", "`cell_fill()`", "`cell_borders()`"),
  Purpose = c("Text styling (color, size, weight, etc.)", "Background color", "Border styling")
) |>
  gt() |>
  fmt_markdown(columns = Function)
```

### Selection helpers (for column selection)

These tidyselect-compatible functions help you specify which columns to target in formatting and styling operations.

```{r}
#| echo: false
dplyr::tibble(
  Function = c("`stub()`", "`row_group()`", "`everything()`", "`starts_with()`", "`ends_with()`", "`contains()`", "`matches()`", "`where()`"),
  Purpose = c("Select stub column", "Select row group column", "Select all columns", "Columns starting with prefix", "Columns ending with suffix", "Columns containing string", "Columns matching regex", "Columns matching predicate")
) |>
  gt() |>
  fmt_markdown(columns = Function)
```

### Utility helpers

These functions provide miscellaneous utilities for table IDs, LaTeX output, and special character handling.

```{r}
#| echo: false
dplyr::tibble(
  Function = c("`random_id()`", "`escape_latex()`", "`gt_latex_dependencies()`"),
  Purpose = c("Generate a random table ID", "Escape special LaTeX characters", "Get LaTeX package requirements")
) |>
  gt() |>
  fmt_markdown(columns = Function)
```
