# Reference and informational functions {#sec-appendix-reference}

```{r setup, include=FALSE, echo=FALSE}
library(gt)
library(dplyr)
```

The **gt** package includes a collection of informational functions that display helpful reference tables. These `info_*()` functions are invaluable when you need to explore available options, discover supported values, or understand formatting styles.

## Date and time style references

### `info_date_style()`

Displays all 41 available date formatting styles for use with `fmt_date()` and `fmt_datetime()`.
 
```{r}
#| eval: false
info_date_style()
```

The resulting table shows:

- **Style number**: The numeric identifier for each style
- **Style name**: The string identifier (e.g., `"iso"`, `"month_day_year"`)
- **Example output**: How each style renders a sample date
- **Flexible**: Whether the style adapts to locale settings

Use style names in your formatting calls:

```r
fmt_date(columns = date_col, date_style = "month_day_year")
fmt_date(columns = date_col, date_style = 14)  
```

### `info_time_style()`

Displays all available time formatting styles for use with `fmt_time()` and `fmt_datetime()`.

```{r}
#| eval: false
info_time_style()
```

The table shows similar information to date styles: style numbers, names, example output, and locale flexibility.

## Currency reference

### `info_currencies()`

Displays a comprehensive table of all supported currencies, including their codes, symbols, and names.

```{r}
#| eval: false
info_currencies()
```

The table includes:

- **Currency code**: The 3-letter ISO 4217 code (e.g., `"USD"`, `"EUR"`, `"JPY"`)
- **Currency symbol**: The symbol used in formatting (e.g., `$`, `€`, `¥`)
- **Currency name**: The full name of the currency

Use currency codes with `fmt_currency()`:

```r
fmt_currency(columns = price, currency = "EUR")
fmt_currency(columns = price, currency = "JPY")
```

## Locale reference

### `info_locales()`

Displays all supported locales for internationalized formatting. Locales affect number formatting, date formatting, and translated text.

```{r}
#| eval: false
info_locales()
```

The table shows locale codes and their corresponding languages/regions. Use these codes with:

- The `locale` argument in `gt()`
- The `locale` argument in formatting functions
- The `locale` argument in `fmt_spelled_num()`

```r
gt(data, locale = "de")
fmt_number(columns = value, locale = "fr")
fmt_date(columns = date, locale = "ja")
```

## Color palette reference

### `info_paletteer()`

The **paletteer** package provides access to hundreds of color palettes from various R packages. The `info_paletteer()` function displays available palettes that work well with `data_color()`.

```{r}
#| eval: false
info_paletteer()
```

The table groups palettes by their source package and shows color swatches for each palette. Use palette names with `data_color()`:

```r
data_color(
  columns = value,
  palette = "viridis::plasma"
)
```

Palette names follow the format `"package::palette"`.

## Flag reference

### `info_flags()`

Displays all available country flags that can be rendered with `fmt_flag()`, along with their country codes.

```{r}
#| eval: false
info_flags()
```

The table shows:

- **Country code (2-letter)**: ISO 3166-1 alpha-2 codes
- **Country code (3-letter)**: ISO 3166-1 alpha-3 codes
- **Country name**: Full country name
- **Flag preview**: The rendered flag icon

Use these codes with `fmt_flag()`:

```r
fmt_flag(columns = country_code)
```

## Icon reference

### `info_icons()`

Displays all available Font Awesome icons for use with `fmt_icon()`.

```{r}
#| eval: false
info_icons()
```

The table shows icon names and rendered previews. Use icon names in your data or with `fmt_icon()`:

```r
dplyr::tibble(
  category = c("file", "folder", "image"),
  icon_name = c("file-alt", "folder", "image")
) |>
  gt() |>
  fmt_icon(columns = icon_name)
```

## Google Fonts reference

### `info_google_fonts()`

Displays recommended Google Fonts for use with **gt** tables, along with information about how to use them.

```{r}
#| eval: false
info_google_fonts()
```

Use Google Fonts with `opt_table_font()` or `google_font()`:
 
```r
gt(data) |>
  opt_table_font(font = google_font("Roboto"))

gt(data) |>
  tab_style(
    style = cell_text(font = google_font("Fira Code")),
    locations = cells_body(columns = code_col)
  )
```

## Unit conversion reference

### `info_unit_conversions()`

Displays all unit conversions supported by the `unit_conversion()` helper function. This is useful when you need to convert between measurement units.

```{r}
#| eval: false
info_unit_conversions()
```

The table shows:

- **Unit class**: The category of measurement (length, mass, volume, etc.)
- **From units**: Available source units
- **To units**: Available target units
- **Conversion factors**: The multipliers used

Use with `unit_conversion()`:

```r
# Get conversion factor from miles to kilometers
unit_conversion(from = "mi", to = "km")

# Apply in a table
dplyr::tibble(distance_miles = c(10, 25, 100)) |>
  gt() |>
  cols_add(distance_km = distance_miles * unit_conversion("mi", "km")) |>
  fmt_number(columns = everything(), decimals = 1)
```

## Built-in datasets

The **gt** package includes 18 datasets for learning and experimentation. These datasets appear throughout the documentation and examples.

### Small example datasets

| Dataset | Description | Rows | Use Case |
|---------|-------------|------|----------|
| `exibble` | A toy tibble with various data types | 8 | Testing formatters |
| `gtcars` | Specifications for 47 luxury/supercars | 47 | Car comparisons |
| `pizzaplace` | A year of pizza sales | 49,574 | Sales analysis |

### Country and population data

| Dataset | Description | Rows | Use Case |
|---------|-------------|------|----------|
| `countrypops` | Country populations 1960-2024 | ~13,000 | Time series, maps |
| `towny` | Ontario municipality populations | 414 | Canadian demographics |
| `peeps` | Personal information for fictional people | 100 | Contact directories |

### Scientific and technical data

| Dataset | Description | Rows | Use Case |
|---------|-------------|------|----------|
| `sza` | Solar zenith angles by location/time | 816 | Heatmaps |
| `constants` | Fundamental physical constants | 30 | Scientific tables |
| `reactions` | Atmospheric reaction rates | 1,344 | Chemistry data |
| `photolysis` | Photolysis rates for organic compounds | 1,590 | Chemistry data |
| `nuclides` | Nuclear data for isotopes | 3,384 | Nuclear chemistry |

### Financial and geographic data

| Dataset | Description | Rows | Use Case |
|---------|-------------|------|----------|
| `sp500` | S&P 500 daily data 1950-2015 | 16,607 | Financial analysis |
| `metro` | Paris Metro stations | 308 | Geographic data |
| `gibraltar` | May 2023 Gibraltar weather | 744 | Weather data |
| `films` | Cannes Film Festival competition | 1,607 | Entertainment data |

### Clinical trial data

| Dataset | Description | Rows | Use Case |
|---------|-------------|------|----------|
| `rx_adsl` | ADSL-style clinical trial data | 182 | Clinical tables |
| `rx_addv` | ADDV-style clinical trial data | 910 | Clinical tables |
| `illness` | Lab tests for illness tracking | 39 | Medical data |

### Using built-in datasets

All datasets are loaded with the package:

```{r}
# View first few rows
head(exibble)

# Use in a table
gtcars |>
  dplyr::select(mfr, model, year, hp, mpg_c) |>
  dplyr::slice(1:5) |>
  gt()
```

## Helper functions quick reference

### Text and content helpers

| Function | Purpose | Example |
|----------|---------|---------|
| `md()` | Interpret text as Markdown | `tab_header(title = md("**Bold Title**"))` |
| `html()` | Interpret text as HTML | `tab_header(title = html("<em>Styled</em>"))` |
| `latex()` | Interpret text as LaTeX | `tab_source_note(latex("\\textit{Note}"))` |

### Unit helpers

| Function | Purpose | Example |
|----------|---------|---------|
| `px()` | Specify pixels | `cols_width(col ~ px(150))` |
| `pct()` | Specify percentage | `cols_width(col ~ pct(25))` |

### Dynamic value helpers

| Function | Purpose | Example |
|----------|---------|---------|
| `from_column()` | Reference column values | `fmt_number(decimals = from_column("precision"))` |
| `currency()` | Custom currency symbol | `fmt_currency(currency = currency(symbol = "BTC"))` |

### Color helpers

| Function | Purpose | Example |
|----------|---------|---------|
| `adjust_luminance()` | Lighten/darken colors | `adjust_luminance("steelblue", steps = 1)` |

### Font helpers

| Function | Purpose | Example |
|----------|---------|---------|
| `google_font()` | Specify Google Font | `opt_table_font(font = google_font("Roboto"))` |
| `default_fonts()` | Get default font stack | `default_fonts()` |
| `system_fonts()` | Get system font stack | `system_fonts("humanist")` |

### Location helpers (for `tab_style()`, `tab_footnote()`, etc.)

| Function | Target |
|----------|--------|
| `cells_title()` | Table title and subtitle |
| `cells_stubhead()` | Stubhead label |
| `cells_column_labels()` | Column labels |
| `cells_column_spanners()` | Column spanner labels |
| `cells_row_groups()` | Row group labels |
| `cells_stub()` | Stub (row labels) |
| `cells_body()` | Body cells |
| `cells_summary()` | Group summary cells |
| `cells_grand_summary()` | Grand summary cells |
| `cells_stub_summary()` | Stub cells in summaries |
| `cells_stub_grand_summary()` | Stub cells in grand summary |
| `cells_footnotes()` | Footnote area |
| `cells_source_notes()` | Source note area |

### Style helpers (for `tab_style()`)

| Function | Purpose |
|----------|---------|
| `cell_text()` | Text styling (color, size, weight, etc.) |
| `cell_fill()` | Background color |
| `cell_borders()` | Border styling |

### Selection helpers (for column selection)

| Function | Purpose |
|----------|---------|
| `stub()` | Select stub column |
| `row_group()` | Select row group column |
| `everything()` | Select all columns |
| `starts_with()` | Columns starting with prefix |
| `ends_with()` | Columns ending with suffix |
| `contains()` | Columns containing string |
| `matches()` | Columns matching regex |
| `where()` | Columns matching predicate |

### Utility helpers

| Function | Purpose |
|----------|---------|
| `random_id()` | Generate random table ID |
| `escape_latex()` | Escape special LaTeX characters |
| `gt_latex_dependencies()` | Get LaTeX package requirements |
